Fail Faster

### 2025-03-24 paper/paper.tex
two

However, careful
measurements show that the performance of existing generator libraries falls far short of what
is possible,
for two reasons.
%
First, the high-level design approach followed by generator combinator
libraries introduces layered abstractions  that are difficult for
compilers to optimize.
Second, the calls a generator makes to its \rand{} can constitute a large
proportion of its run-time, magnifying the cost of an inefficient
implementation.
%
In this paper, we characterize, quantify, and address these issues.

@@

We isolate and quantify the performance benefits of faster randomness, showing
that a DSL's choice of \rand{} has a dramatic effect on its performance. Based
on the observation that the OCaml implementation of \bq{}'s \rand{} is slow in
a way that is simple to fix, we perform a controlled experiment, comparing fast and slow
versions of the same library to analyze the impact of faster \rands{} on
generator performance.

@@

Last, many PBT libraries---including both \bq and ScalaCheck---implement their \texttt{weighted\_union} combinators
in ways that are asymptotically more efficient but slower in common cases than a more naive algorithm.
Weighted union uses the Fitness Proportionate Selection algorithm~\cite{geneticalgorithm}, which (1) samples a number $r$
between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
a binary search. However, since the lists are short in practice, a linear scan is almost always faster.
Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
perform this search, which incurs further run time overhead.

@@

\subsection{Let-Insertion and Effect Ordering}
Careful readers might note that the definition of \texttt{bind} (Figure~\ref{fig:bind-with-let-binding})
was more complicated than one might expect. Why not define bind in the standard way
for a reader monad (Figure~\ref{fig:bind-without-let-binding})?
Unfortunately, the latter definition is wrong in our context as it leads to incorrect code being generated.
For example, consider \texttt{Gen.bind' (Gen.int .<0>. .<1>.) (fun x -> Gen.return .<(.~x,.~x)>.)}.
This generates the run time code \texttt{fun size random -> (SR.int random 0 1, int SR.random 0 1)},
which is incorrect.  Instead of generating a single integer and returning it
twice, it samples two different integers.
This matters because, as described in Section~\ref{section:intro}, \camlname and
\scalaname are intended to be equivalent to their unstaged counterparts.

@@

Our implementation in OCaml uses the PPX (PreProcessor eXtension) system to synthesize
staged generators from type definitions. However, any language that supports type-derived
generators can implement a staged version using a similar implementation to the original. In Scala,
for instance, the same strategy could be realized using typeclass
resolution. In both settings, the result is a three-stage process: a
metaprogram---via PPX or typeclasses---constructs a generator expression
composed of staged combinators; this expression is evaluated at compile
time to produce a specialized generator; and finally, the generator is
executed at runtime to produce values.

@@

We run each
strategy on all tasks using a 60-second timeout.  If a bug is not found within
this limit, the task is considered ``unsolved.''  We exclude tasks where all strategies fail to
find the bug from our dataset, as they provide no basis for
comparison.  Similarly, we exclude tasks where the \bq strategy completes in
under 5ms, as such negligible runtimes do not yield meaningful insights into
relative performance. Across all 30 seeds, these filters remove
28/600 (4.67\%) of type-derived STLC's tasks,
168/600 (28\%) of STLC's tasks,
149/1110 (14.42\%) of type-derived BST's tasks,
268/1110 (24.1\%) of single-pass BST's tasks, and
371/1110 (33.42\%) of insertion-based BST's tasks. More sophisticated
strategies tend to find bugs very quickly, leading to a higher number of filtered
tasks. By applying these filters, we ensure that our reported speedups reflect
optimizations that meaningfully impact performance.

>>>>

However, careful
measurements show that the performance of existing generator libraries falls far short of what
is possible,
for two reasons.
%
First, the high-level design approach followed by generator combinator
libraries introduces layered abstractions  that are difficult for
compilers to optimize.
Second, the calls a generator makes to its \rand{} can constitute a large
proportion of its run time, magnifying the cost of an inefficient
implementation.
%
In this paper, we characterize, quantify, and address these issues.

@@

We isolate and quantify the performance benefits of faster randomness, showing
that a DSL's choice of \rand{} has a dramatic effect on its performance. Based
on the observation that the OCaml implementation of \bq{}'s \rand{} is slow in
a way that is simple to fix, we perform a controlled experiment, comparing fast and slow
versions of the library to analyze the impact of faster \rands{} on
generator performance.

@@

Last, many PBT libraries---including both \bq and ScalaCheck---implement their \texttt{weighted\_union} combinators
in ways that are asymptotically more efficient but slower in common cases than a more naive algorithm.
Weighted union uses the Fitness Proportionate Selection algorithm~\cite{geneticalgorithm}, which (1) samples a number $r$
between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
a binary search. However, since the lists are short in practice, a linear scan is almost always faster.
Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
perform this search, which incurs further run-time overhead.

@@

\subsection{Let-Insertion and Effect Ordering}
Careful readers might note that the definition of \texttt{bind} (Figure~\ref{fig:bind-with-let-binding})
was more complicated than one might expect. Why not define bind in the standard way
for a reader monad (Figure~\ref{fig:bind-without-let-binding})?
Unfortunately, the latter definition is wrong in our context as it leads to incorrect code being generated.
For example, consider \texttt{Gen.bind' (Gen.int .<0>. .<1>.) (fun x -> Gen.return .<(.~x,.~x)>.)}.
This generates the run-time code \texttt{fun size random -> (SR.int random 0 1, int SR.random 0 1)},
which is incorrect.  Instead of generating a single integer and returning it
twice, it samples two different integers.
This matters because, as described in Section~\ref{section:intro}, \camlname and
\scalaname are intended to be equivalent to their unstaged counterparts.

@@

Our implementation in OCaml uses the PPX (PreProcessor eXtension) system to synthesize
staged generators from type definitions. However, any language that supports type-derived
generators can implement a staged version using a similar implementation to the original. In Scala,
for instance, the same strategy could be realized using typeclass
resolution. In both settings, the result is a three-stage process: a
metaprogram---via PPX or typeclasses---constructs a generator expression
composed of staged combinators; this expression is evaluated at compile
time to produce a specialized generator; and finally, the generator is
executed at run time to produce values.

@@

We run each
strategy on all tasks using a 60-second timeout.  If a bug is not found within
this limit, the task is considered ``unsolved.''  We exclude tasks where all strategies fail to
find the bug from our dataset, as they provide no basis for
comparison.  Similarly, we exclude tasks where the \bq strategy completes in
under 5ms, as such negligible run times do not yield meaningful insights into
relative performance. Across all 30 seeds, these filters remove
28/600 (4.67\%) of type-derived STLC's tasks,
168/600 (28\%) of STLC's tasks,
149/1110 (14.42\%) of type-derived BST's tasks,
268/1110 (24.1\%) of single-pass BST's tasks, and
371/1110 (33.42\%) of insertion-based BST's tasks. More sophisticated
strategies tend to find bugs very quickly, leading to a higher number of filtered
tasks. By applying these filters, we ensure that our reported speedups reflect
optimizations that meaningfully impact performance.
